---
title: "BIOL-607 Biological Data Analysis Homework 1: Sampling and Simulation"
author: "Andrew Judell-Halfpenny"
date: "September 21, 2018"
output: pdf_document
---

# 1. R Warmup
## 1a. Create a vector of 100 randomly distributed numbers between 0 and 100 using runif and save the vector into the variable my_vec. If you cannot figure out runif, try ?runif. What information does str and summary tell you about my_vec? How do they differ?
```{r}
# library initialization
library(dplyr)
library(purrr)
library(magrittr)
#random vector
my_vec<-runif(100, min=0, max=100)
class(my_vec)
str(my_vec)
summary(my_vec)
```
## 1a. Answer:
### The str() function returns the class of the object, its dimensions, and returns the first few entries in the object.

## 1b. Try out a little plot. What does hist(my_vec) show you? What information does the helpfile for hist tell you about what you just did?
```{r}
hist(my_vec)
??hist

```
## 1b. Answer:
### The hist() function produces a simple histogram of an object.  The helpfile produces a deluge of information about the type of data to which the function can be applied as well as the arguments that are expected by the functon.

## 1c. Load up the mtcars data set using data(mtcars). Apply the following functions to mtcars: class, str, summary. What do these three functions tell you about mtcars?
```{r}
data(mtcars)
class(mtcars)
str(mtcars)
summary(mtcars)
```

## 1c. Answer:
### These functions describe the mtcars object as a data frame with 32 rows and 11 named columns of numeric data.  The summary() function produces a five number summary for each column in the data frame that includes the minimum, mean, median, maximum, first quartile, and third quartile values.

## 1d. Look at the help file for the class data.frame. What does it tell you about these objects?
```{r}
??dplyr::data.frame
```
## 1d. Answer:
### These files are not as strongly typed as matrices or numerics.  They can accept strings, floats, integers, and even multiple values as entries within a single data frame object.  Data frames can also include multiple classes of data in a single object.

# 2. Sample Properties
## 2a. Produce descriptive statistics for the following vasopressin expression levels in voles.
```{r}
vole_vaso <- c(98,96,94,88,86,82,77,74,70,60,
           59,52,50,47,40,35,29,13,6,5)

```


## 2b. What is the mean, median, sd, and interquartile range of the sample
```{r}
mean(vole_vaso)
median(vole_vaso)
sd(vole_vaso)
IQR(vole_vaso)

```

## 2c. What is the standard error of the mean (do this with a formula!)?
```{r}
vole_vaso
std.er <- function(x){
  output <-sd(x)/sqrt(length(x))
  return(output)}
std.er(vole_vaso)
sd(vole_vaso)
```

## 2d. What does the standard error of the mean tell you about our estimate of the mean values of the population of vole vassopressin?
## 2d. Answer:
### The standard error of the mean is a measurement of the precision of the estimate of the mean of a population for a particular sample size of that population.

# 3. Sample Size for upper quartiles.

## Let’s assume this sample of vole vassopressin expression is representative of the popultion.    
## We can get the upper quartile value of vole vassopressin with
```{r}
quantile(vole_vaso, probs = 0.75)

```

## 3a. Use sample() to get just one resample with a sample size of 10. What is its upper quartile?
```{r}
vole_vaso.resamped<-sample(x=vole_vaso,10, replace=T)
quantile(vole_vaso.resamped, probs=0.75)
```

## 3b. Build an initial data frame for simulations with the sample sizes 5 through 20. Have 100 simulations per sample size.
```{r}
# initialize data frame description
for (i in 5:20){
   sampSim <- data.frame(samp_size=c(5:20),sim_num=list(100))}
str(sampSim)
# initialize data frame of appropriate size
for (i in 5:20){
   sampSim[[i]]<-map(i, ~data.frame(samp_size=rep(i,times=100)))}
str(sampSim)

x<-list()
for (i in 5:20){
   x[[i]]<-map(1:100, ~map(i, ~sample(vole_vaso, i,replace=T)))
   names(x[[i]])<-replicate(100, expr=paste0("samp_size",sep="",i))}
str(x)

```


### 3c. Use this data frame to get simulated upper quartiles for each sample size.
```{r}
for (i in 5:20){
   print(map(i, ~quantile(sample(vole_vaso, i,replace=T), probs=0.75)))}

```

### 3d. With a plot, make a guesstimate as to the best sample size for estimating the upper quartile of the population.
```{r}
for (i in 5:20){
   print(map(i, ~std.er(sample(vole_vaso, i,replace=T))))}


for (i in 5:20){
   qnts<-map(i, ~quantile(sample(vole_vaso, i,replace=T), probs=0.75))
   qnts.df<-data.frame(qnts)
   qnts.df
   print(qnts.df)
   std.er(qnts.df)
   sd(qnts.df)}

   
str(qnts.df)


```


# 4. SE and CI

## 4a. With the upper quartile simulations, calculate the SD for each sample size using dplyr
## 4b. What does this value, the standard error of the upper quartile, mean?
## 4c. What is the CI of the upper quartile with a sample size of 10. What does this mean?

```{r}
sample.size<-c()
vole_vaso.df<-data.frame(vole_vaso)
map(5:20 , ~sample.size)
ss<-apply(vole_vaso,FUN=sample)
samp_sim<-map(5:20. ~sample(1:nrow(vole_vaso.df),size=map(5:20) replace=T))
head(samp_sim)
for(i in 5:20){
   print(i)
   samp_sim.df<-data.frame<-map(1:100, ~map(5:20, ~sample(vole_vaso, size=i, replace=T)))}
   print(samp_sim.df)
for (i in 5:20){
   print(i)
   map(1:100, ~map(1:length(c(5:20)),~sample(vole_vaso, size=i, replace=T))) %>%
   str()}
```

## 4d. Extra Credit: Instead of having each sim calculate a upper quartile, and then sifting down to the SE of that upper quartile (or the CI), have each simulation calculate a CI. Then, for just a sample size of 10, how many CI’s contain the true value of 83?
